---
title: VRC01 neutralization report
author: David Benkeser
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    highlight: haddock
---

# Exploring markdown

Markdown is a way of writing plain text that, upon rendering, looks formatted.

For example, we can __bold text__ or *italicize* it.

We can also ~~strikethrough text~~.

Sometimes `monospace font` is nice for formatting as well.

## Subheadings are fun too

### Even smaller subheadings are also possible

We can create bullet points:

* that is pretty helpful
* for listing things

We can create numbered lists:

1. item number 1
2. item number 2

We can also [include hyperlinks](www.emory.edu).

# Including R code

There are two options for inputting code in R Markdown.

## Code chunks

```{r, name-of-chunk}
print("hello from R!")
```

```{r, read-data}
# replace with your own
abs_file_path <- "~/Dropbox/Emory/Teaching/DSTK/intro_repro_workflo/lectures/04_rmarkdown/"
full_file_path <- paste0(
  abs_file_path, "hiv_project/data/vrc01_data.csv"
)
data <- read.csv(full_file_path, header = TRUE)
head(data)
```

## Inline R expressions

We can use inline `R` expressions to report numbers in the text. For example, `vrc01_data.csv` has `r nrow(data)` rows and `r ncol(data)` columns.
