---
title: VRC01 neutralization report
author: David Benkeser
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: html_document
params:
  production: FALSE
  cutpoint: 1
---

```{r, chunk-opts, include = FALSE, message = FALSE, warning = FALSE, eval = params$production}
# if production == FALSE, then this code will not
# be evaluated and thus code chunks will appear in 
# the rendered document
library(knitr)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE
)
```


```{r, load-data, echo = FALSE}
# replace with your own
abs_file_path <- "~/Dropbox/Emory/Teaching/DSTK/intro_repro_workflo/lectures/04_rmarkdown/"
full_file_path <- paste0(
  abs_file_path, "hiv_project/data/vrc01_data.csv"
)
data <- read.csv(full_file_path, header = TRUE)
```

# Descriptive Analysis

## Tabular analysis

Characteristics of the `nrow(data)` viruses from the CATNAP database are displayed in the table below.

```{r, table1}
# labelled + gtsummary package for making pretty html tables
library(labelled)
library(gtsummary)

# label variables
var_label(data) <- list(
  id = "ID",
  ab_resistance = "Antibody resistance",
  shield_glycans = "Shield glycans",
  region = "Region",
  env_length = "Length of Env protein"
)

data$number_glycans <- ifelse(data$shield_glycans < 4, "< 4", ">= 4")

# make the table
data |> 
  select("region", "env_length", "ab_resistance", "number_glycans") |>
  tbl_summary(by = number_glycans) |> 
  modify_spanning_header(c("stat_1", "stat_2") ~ "**PNGs in Glycan Shield**") |>
  add_overall() |>
  add_p()
```


## Graphical analysis

Below we report a scatter plot of resistance as a function of PNGs in the glycan shield.

```{r, plot-data}
#| fig.cap = "Scatter plot with linear model smoother",
#| fig.align = "center", 
#| out.width = "600px"

library(ggplot2)
ggplot(data, aes(x = shield_glycans, y = ab_resistance)) + 
  geom_point() +
  geom_smooth(method = lm) + 
  theme_bw()
```

# Regression Analysis

## Primary model

```{r, fit-primary-model}
mod <- glm(
  ab_resistance ~ shield_glycans + region + env_length,
  data = data
)

tbl_regression(mod) |>
  add_global_p()
```

## Secondary model

In this analysis, we analyzed a binary endpoint where level 1 means antibody resistance score was greater than `params$cutpoint`.

```{r, binary-model}
binary_mod <- glm(
  I(ab_resistance > params$cutpoint) ~ shield_glycans + region + env_length,
  data = data, 
  family = binomial()
)

tbl_regression(binary_mod, exponentiate = TRUE) |>
  add_global_p()
```