---
title: VRC01 neutralization report
author: David Benkeser
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output: html_document
params:
  production: FALSE
---

```{r, chunk-opts, include = FALSE, message = FALSE, warning = FALSE, eval = params$production}
library(knitr)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE
)
```

```{r, load-data, echo = FALSE}
data <- readRDS(here("output", "data_clean.rds"))
```

# Descriptive Analysis

## Tabular analysis

Characteristics of the `r nrow(data)` viruses from the CATNAP database are displayed in the table below.

```{r, table1}
table_one <- readRDS(here("output", "table1.rds"))
table_one
```

## Graphical analysis

Below we report a scatter plot of resistance as a function of PNGs in the glycan shield.

```{r, plot-data}
#| fig.cap = "Scatter plot with linear model smoother",
#| fig.align = "center", 
#| out.width = "600px"
knitr::include_graphics(
  here("figs", "scatter.png")
)
```

# Regression Analysis

```{r, load-models}
models <- readRDS(here("output", "models.rds"))
```

## Primary model

```{r, primary-model}
library(gtsummary)
tbl_regression(models$primary) |>
  add_global_p()
```

## Secondary model

In this analysis, we analyzed a binary endpoint where level 1 means antibody resistance score was greater than `r models$secondary_cutpoint`.

```{r, binary-model}
tbl_regression(models$secondary, exponentiate = TRUE) |>
  add_global_p()
```