---
title: "Project organization"
author: "David Benkeser, PhD MPH<br> <span style = 'font-size: 50%;'> Emory University<br> Department of Biostatistics and Bioinformatics </span>"
date: 'Introduction Reproducible Workflows in R<br><br>`r renv::activate("../.."); fontawesome::fa("link", fill = '#ffffff')` [.white[bit.ly/intro_repro_workflo]](https://bit.ly/intro_repro_workflo) <br> `r fontawesome::fa("book", fill = '#ffffff')` [.white[Additional reading]](https://benkeser.github.io/intro_repro_workflo/readings#project-organization)'
tldr: "Choose an organization structure that sparks joy."
recording: ""
start: "0:0:0"
reading: "project-organization"
breakoutsoln: ""
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      titleSlideClass: ["left", "middle", "inverse"]
      highlightStyle: tomorrow-night-blue
      highlightLanguage: rmarkdown
      highlightLines: false
---

background-color: #007dba
class: title-slide, center, inverse, middle

*File organization and naming are powerful weapons against chaos.*
<div> [Jenny Bryan](https://jennybryan.org/)

???

Project organization may seem a rather dull/pedantic topic, but it is actually very important!

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)

extra_css <- list(
  ".small" =  list(`font-size` = "80%"),
  ".large" =  list(`font-size` = "150%"),
  "ul li" = list(`margin-bottom` = "10px"),
  ".red" = list(color = "#da291c",
                `font-weight` = "600"),
  ".green" = list(color = "#348338",
                `font-weight` = "600"),
  ".usera" = list(color = "#006c5b", `font-weight` = "600"),
  ".userb" = list(color = "#6d2077", `font-weight` = "600"),
  ".code-box" = list(display = "inline-block", 
    width = "95%",
    `line-height` = "1em",
    padding = "5px",
    `font-size` = "80%",
    `background-color` = "#f6f6f6",
    `align-items` = "left",
    `font-family` = "Courier Prime",
    border = "1px solid black"),
  "ul" = list(`margin-bottom` = "0.5em",
              `margin-top` = "0.5em"),
  "p" = list(`margin-bottom` = "0.25em")
)

style_mono_accent(
  base_color = "#012169",
  header_font_google = google_font("DM Sans"),
  text_font_google   = google_font("DM Sans", "400", "400i"),
  code_font_google   = google_font("Courier Prime"),
  extra_css = extra_css
)
```

<style type="text/css">
.remark-slide-content {
    font-size: 25px
}
</style>


---

### Basic principles

__Major suggestions__
* Be consistent across projects.
* Raw data are sacred. Keep them separate from everything else.
* Separate code and data.
* .red[Put everything in one directory.]
* .red[No absolute filepaths in code.]
* Modularize code.

__Other suggestions__
* `Makefile`'s and/or READMEs to document dependencies.
* No spaces in file names. 
* Use meaningful file names.
* Use YYYY-MM-DD date formatting.
* Use a package management system.

???

There is not one correct way to organize a project, but there are many incorrect ways. Over time you will develop your own system for doing so. We'll discuss a few frameworks that may be helpful. 

The biggest thing is to commit to using the same structure for as long as its working for you. As you are working on many projects, be self-aware during the process. Are you consistently losing track of certain types of files? Are there gaps in the workflow that make things hard to reproduce? What can you do to fill in the gaps?

---

background-color: #84754e
class: title-slide, center, inverse, middle

*You mostly collaborate with yourself, and me-from-two-months-ago never responds to email.*
<div> [Karen Cranston](http://kcranston.github.io/)

???

This is as much for preserving your own sanity as it is about being reproducible (though it is about that too). There's a (big) upfront cost here, but in the end it is worth it. 

---

### What to organize?

Depending on the nature of the project, the organization system may look different.
* E.g., data analysis vs. methods paper vs. presentation vs. web app vs. etc...

The systems should adhere to the same general principles, but different requirements may necessitate different structures. 

.red[Think about organization of a project from the outset!]

???

Thinking about project organization from the outset is the most important thing. Map out (e.g., as comments in a README) how you see the project developing. Even if things change over time, it's good to have a structure in place from the beginning. 

This pertains to project directories in particular, but also to your hard drive in general. I made the mistake of organizing projects by institution (UW, Berkeley, Emory), but am now regretting that. What about projects that weren't finished when I got to Emory? What about projects that I continued developing once I got here. It's a mess. Don't be like me. 

---

### Organizing data

Raw data are sacred... but may be a mess. 
* You'll be surprised (and disheartened) by how many color-coded excel sheets you'll get in your life.

Tempting to edit raw data by hand. .red[Don't!]
* Everything scripted!

Use meta-data files to describe raw and cleaned data.
* structure as data (e.g., `.csv`) so easy to read

???

The generation of the raw data may be the one thing out of your control in an analysis. But from the time the data are passed on to you, everything that happens should be reproducible. 

This can be painful. For many projects, 90% of time might be devoted to wrangling raw data into a format that is usable. Save yourself from the danger of having to re-do all those painful bits when (not if) the data change.

---

### Organizing data

Hadley Wickham [defined the notion of tidy data](https://doi.org/10.18637/jss.v059.i10). 
* Each variable forms a column.
* Each observation forms a row.
* Each observational unit forms a table. 

.code-box[
| ptid | day | age | drug | out |
|------|-----|-----|------|-----|
| 1    | 1   | 28  | 0    | 0   |
| 1    | 2   | 28  | 0    | 1   |
| 2    | 1   | 65  | 0    | 0   |
| 2    | 2   | 65  | 1    | 1   |
| 3    | 1   | 34  | 0    | 0   |
| 3    | 2   | 34  | -    | 1   |
]

---

background-color: #487f84
class: title-slide, center, inverse, middle

*If the first line of your R script is*

`setwd("C:\Users\jenny\path\that\only\I\have")`

*I will come into your office and SET YOUR COMPUTER ON FIRE 🔥.*


<div> [Jenny Bryan](https://jennybryan.org/)

---

### Project-based organization

Organize each analysis into a separate *project*.
* A directory on your computer that holds all relevant files

The __project directory__ is the directory that holds all relevant files.

Any R script in the project assumes it will be run: 
* from a fresh R process 
* with working directory set to the project directory

Read [Jenny Bryan and James Hester's chapter](https://rstats.wtf/project-oriented-workflow.html) on project-oriented work-flows.

---

### The `here` package

.red[No absolute paths.]

For `R` projects, the [`here`](https://here.r-lib.org/) package provides a simple way to use relative file paths.

Each `R` script or `Rmd` report, should contain a call to `here::i_am('path/to/this/file')` at the top.
* `path/to/this/file` should be replaced with the path .green[relative] to the __project directory__.
* `here::i_am` means use function `i_am` from `here` package.

---

class: title-slide, center, inverse, middle
background-color: #487f84

.large[
`r fontawesome::fa("desktop", fill = "#ffffff")` .white[Updating report with here package]
]

.large[
`r fontawesome::fa("code", fill = "#ffffff")` <a href="01_hiv_project.zip" style="color:#ffffff" download>Code: `01_hiv_project.zip`</a>
]

???

Demo: 
* `01_hiv_project.zip`
* Modify `hiv_report.Rmd`'s chunk labeled `load-data` to include call to `i_am`
* Update filepath accordingly
* Render document

---

### The `here` package

The call to `i_am` establishes the __project directory__.
* Subsequent file paths can be made using the `here` function.
* `here("path", "to", "folder", "file.csv)` = `[project_directory]/path/to/folder/file.csv`

You can check what `here` thinks the root directory is by printing the output of `here::here()`.

---

### Modularize code

It is common, but not best practice, to maintain all `R` code in an `Rmd` file.
* OK for small projects, but difficult to maintain

Reasons to prefer a different system:
* Faster compiling of reports
  * only update portions of report that needs updating
* Easier to locate chunks of code
* More likely to write code that can be re-used

???

R Markdown has a cache system that makes report compiling faster. However, it is somewhat rudimentary and may be considered better practice to develop your own caching system, e.g., based on `Make`. 

---

### Modularize code

Separate R data analysis into discrete tasks and have a separate R script for each.
* cleaning data
  * `code/00_clean_data.R`
* descriptive analysis
  * `code/01_table1.R`
  * `code/02_scatter.R`
* primary model
  * `code/03_models.R`

---

class: title-slide, center, inverse, middle
background-color: #487f84

.large[
`r fontawesome::fa("desktop", fill = "#ffffff")` .white[Modularizing code]
]

.large[
`r fontawesome::fa("code", fill = "#ffffff")` <a href="02_hiv_project.zip" style="color:#ffffff" download>Code: `02_hiv_project.zip`</a>
]

???

Demo: 
* `02_hiv_project.zip`
* rename `data` to `raw_data`
* make `output` and `code` folder
* remove pieces of `hiv_report.Rmd` to `R` files
* remove `cutpoint` from `params` and instead feed arguments into script from command line

---

### The `renv` package

.red[Use a package management system.]

To increase reproducibility of a project, we must keep track of what packages are used.
* Want to avoid chasing down 100 errors like this:

.code-box[
.red[Error in library(ggplot2) : there is no package called ‘ggplot2’] <br>
]

The [`renv` package](https://rstudio.github.io/renv/articles/renv.html) is useful to this end.

---

### The `renv` package

Download the <a href='example_project.zip' download>`example_project`</a> folder.

.code-box[
example_project/ <br>
&nbsp;&nbsp;&nbsp; Makefile <br>
&nbsp;&nbsp;&nbsp; figs/ <br>
&nbsp;&nbsp;&nbsp; R/ <br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; R/barchart.R <br>
&nbsp;&nbsp;&nbsp; Rmd/ <br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Rmd/report.Rmd <br>
]

If you scan the code, you will see we need __three R packages__.
* `here`, `wesanderson`, `knitr`, and `rmarkdown`

---

### The `renv` package

Open an `R` session from the `example_project` folder. 
* Install the `renv` package if needed.

Initialize the project by running the following command.

```{r, eval = FALSE}
renv::init()
```

You will see a lot of output. What just happened??

---

### `renv.lock` file

You should now see __lockfile__, `example_project/renv.lock`.

````
{
  "R": {
    "Version": "4.0.3",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://cloud.r-project.org"
      }
    ]
  },
  "Packages": {
    "base64enc": {
      "Package": "base64enc",
      "Version": "0.1-3",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "543776ae6848fde2f48ff3816d0628bc"
    },
  ...
````

---

### `renv.lock` file

This __lockfile__ records all of the information about packages needed by your project.
* Version of package, where was it installed from, etc...

How does it know?
* `renv` scans all files in your project directory.
* Looks for `library`, `require`, or `package::function`.

---

### `.Rprofile`

We also see `example_project/.Rprofile`, containing one line.

```{r, eval = FALSE}
source("renv/activate.R")
```

We have not talked about `.Rprofile`'s because they are generally antithetical to reproducible research.

When `R` starts, it searchers for `.Rprofile` and runs what it finds.
* Use this to change various options, always load packages, etc...

In this case, whenever we start `R` in `example_project`, we __activate__ our `R` environment.
* Telling `R` where packages for this project are saved, etc...
* Details are not too important.

---

### `renv` folder

You will also see a folder `example_project/renv/`.
* This folder contains your __project library__.

`renv` tries to be clever about installing packages.
* Already have a package installed elsewhere? `renv` will link to it.
* Otherwise, package is installed in `renv/library`.

Note that `renv/.gitignore` ensures packages not put under version control.

---

### Activating `renv`

`renv` will automatically be active in any `R` session that is run from the `example_project` directory.
* Recall the presence and function of `.Rprofile`.

To activate in an `R` session run from elsewhere: `renv::activate('path/to/renv')`.

For using `renv` with R Markdown projects, this is important!
* Either need to `activate` in a code chunk.
* Or use `knitr::opts_knit$set(root.dir = here::here())` (assuming the project root contains your `renv` project library).

---

### Collaborating with `renv`

A typical `renv` collaborative workflow on GitHub:

* .usera[User A] initializes the lockfile using `renv::init()`.
* .usera[User A] commits `renv.lock`, `.Rprofile`, and `renv/activate.R` and pushes to GitHub.
* .userb[User B] pulls from GitHub, opens `R`, and uses `renv::restore()` to synchronize their local project directory.
* .userb[User B] adds new packages to code, uses `renv::snapshot()` to record changes to `renv.lock`
* .userb[User B] commits `renv.lock` and pushes to GitHub.
* .usera[User A] pulls from GitHub, opens `R`, and uses `renv::restore()` to synchronize their local project directory.
* ...

---

### Breakout exercise

With a partner, choose .usera[User A] and .userb[User B].

.usera[User A] 
* initialize the lockfile for `example_project`
  * open `R` in `example_project` and run `renv::init()`
* use `git init` to initialize version control of `example_project`
* create a GitHub repository for `example_project`
* add the GitHub repository as a remote to your local repository
  * .small[`git remote add origin https://github.com/usera/example_project.git`]
  * .small[`git remote add origin git@github.com:usera/example_project.git`]
* commit all files locally and push to repository

---

### Breakout exercise

.userb[User B]
* create a fork of .usera[User A]'s repository
* `git clone` the fork to your local machine
* `cd` into `example_project` and open `R`
* run `renv::restore()` to synchronize package library
* confirm that you can build the report
  * run `make report` from `example_project` directory
  * open `Rmd/report.html` to confirm correct build

---

### Breakout exercise

.userb[User B] now wants to change the colors of the graph
* open `R` in `example_project` directory
* run `renv::remove('wesanderson')` to remove the `wesanderson` package from the lockfile
* replace lines 4-5 of `barchart.R` with the following and save

```{r, eval = FALSE}
library(RColorBrewer)
colors <- brewer.pal(3, "Dark2")
```

* open `R` in `example_project` and run `renv::status()`
* if prompted, run `install.packages("RColorBrewer")` 
* run `renv::snapshot()` to add `RColorBrewer` to the lockfile
* commit changes and push to your fork
* submit a PR to .usera[User A]'s repository

---

### Breakout exercise

.usera[User A]
* add a remote linking to user B's repository
  * .small[`git remote add userb https://github.com/userb/example_project.git`]
* `fetch` .userb[User B]'s `master` branch
  * .small[`git fetch userb master`]
* `checkout` .userb[User B]'s `master` branch
  * .small[`git checkout remotes/userb/master`]

---

### Breakout exercise

.usera[User A]

* `cd` into `example_project` and open `R`
* run `renv::restore()` to synchronize your package library
* confirm that you can build the report
  * run `make report` from `example_project` directory
  * open `Rmd/report.html` to confirm correct build

---

### Breakout exercise

.usera[User A]

* if the report builds correctly, create a local branch named `userb` from .userb[User B]'s `master` branch
  * .small[`git checkout -b userb`]
* `merge` .userb[User B]'s `master` into your `master` 
  * .small[`git checkout master`]
  * .small[`git merge userb`]
* close the PR by pushing to GitHub
  * .small[`git push origin master`]

---

### Breakout exercise

.usera[User A] has changed their mind about the colors!

* open `R` in `example_project` directory
* run `renv::remove('RColorBrewer')` to remove `RColorBrewer` from the lockfile
* replace lines 4-5 of `barchart.R` with the following and save

```{r, eval = FALSE}
colors <- c("red", "blue", "green")
```
* open `R` and run `renv::status()` and `renv::snapshot()`
* confirm that you can build the report
  * run `make report` from `example_project` directory
  * open `Rmd/report.html` to confirm correct build
* if the report builds correctly, `commit` and `push`

---

### Breakout exercise

.userb[User B]

* add a remote linking to user A's repository
  * .small[`git remote add usera https://github.com/usera/example_project.git`]
* `fetch` .usera[User A]'s `master` branch
  * .small[`git fetch usera master`]
* `checkout` .usera[User A]'s `master` branch
  * .small[`git checkout remotes/usera/master`]

---

### Breakout exercise

.userb[User B]

* `cd` into `example_project` and open `R`
* run `renv::status()` and, if needed, `renv::restore()` to synchronize your package library
* confirm that you can build the report
  * run `make report` from `example_project` directory
  * open `Rmd/report.html` to confirm correct build

---

### Breakout exercise

.userb[User B]

* if the report builds correctly, create a local branch named `usera` from .usera[User A]'s `master` branch
  * .small[`git checkout -b usera`]
* `merge` .usera[User A]'s `master` into your `master` 
  * .small[`git checkout master`]
  * .small[`git merge usera`]
* update your fork of the repository by pushing to GitHub
  * .small[`git push origin master`]
