---
title: "Package management"
author: "David Benkeser, PhD MPH<br> <span style = 'font-size: 50%;'> Emory University<br> Department of Biostatistics and Bioinformatics </span>"
date: 'Introduction Reproducible Workflows in R<br><br>`r renv::activate("../.."); fontawesome::fa("link", fill = '#ffffff')` [.white[bit.ly/intro_repro_workflo]](https://bit.ly/intro_repro_workflo) <br> `r fontawesome::fa("book", fill = '#ffffff')` [.white[Additional reading]](https://benkeser.github.io/intro_repro_workflo/readings#project-organization)'
tldr: "renv holds the title of 'least-bad' package management system for R."
recording: ""
start: ""
reading: "package-management"
breakoutsoln: ""
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      titleSlideClass: ["left", "middle", "inverse"]
      highlightStyle: tomorrow-night-blue
      highlightLanguage: rmarkdown
      highlightLines: false
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)

extra_css <- list(
  ".clearbox" = list(
    display = "inline-block", 
    width = "100%",
    padding = "5px",
    border = "1px solid #012169",
    `background-color` = "#d9d9d6",
    `align-items` = "left",
    `vertical-align` = "middle"
  ),
  ".small" =  list(`font-size` = "80%"),
  ".large" =  list(`font-size` = "150%"),
  "ul li" = list(`margin-bottom` = "10px"),
  ".red" = list(color = "#da291c",
                `font-weight` = "600"),
  ".green" = list(color = "#348338",
                `font-weight` = "600"),
  ".remark-code-line-highlighted" = list(border = "2px solid #f2a900",
                                         `background-color` = "#002451"),
  "a" = list(color = "#007dba"),
  "a:hover" = list("text-decoration" = "underline")
)

style_mono_accent(
  base_color = "#012169",
  header_font_google = google_font("DM Sans"),
  text_font_google   = google_font("DM Sans", "400", "400i"),
  code_font_google   = google_font("Courier Prime"),
  extra_css = extra_css
)
```
<style type="text/css">
.remark-slide-content {
    font-size: 22px
}
</style>

## Learning objectives

1. Identify the benefits and limitations of package management systems in `R`.
2. Describe the basic workflow of using `renv`.

---

## Learning objectives: redux

1. Describe the role of the `renv.lock` file.
2. Describe the role of `renv/activate.R`.
3. Initialize, restore, and take snapshots of a project library.
4. Use `.libPaths()` to identify library locations during `R` run time.

---

### Reproducibility: redux

Thus far, we have mainly focused on making project reproducible on a limited basis.
* Single system, single `R` installation

However, we would ideally like to make projects work on different systems.
* Good: my work PC + my laptop
* Better: my work PC + my laptop + your laptop
* Best: anywhere!

This chapter will take us a step closer to Better via __package management__.

???

Containerization (e.g., Docker) is really the only way to achieve the Best level.

---

### The `renv` package

The `renv` workflow

1. `renv::init()` to initialize a new project-specific environment
2. work on the project as normal, installing/removing R packages as needed
3. `renv::snapshot()` to save the state of the project library to `renv.lock`

Having a project-specific environment allows you to:
* record which (version of which) packages are used
	* and where those packages were downloaded
* have different versions of the same package on your computer without confusing `R`
* synchronize library across different computers

---

class: title-slide, center, inverse, middle
background-color: #487f84

.large[
`r fontawesome::fa("desktop", fill = "#ffffff")` .white[Initializing `renv` in HIV project]
]

.large[
`r fontawesome::fa("code", fill = "#ffffff")` <a href="01_renv_hiv_project.zip" style="color:#ffffff" download>Code: `01_renv_hiv_project.zip`</a>
]

???

Demo: 
* `01_renv_hiv_project.zip`
* run `renv::init()` and examine output

---

### `renv.lock` file

You should now see __lockfile__, `renv.lock`

````
{
  "R": {
    "Version": "4.1.0",
    "Repositories": [
      {
        "Name": "CRAN",
        "URL": "https://mirrors.nics.utk.edu/cran"
      }
    ]
  },
  "Packages": {
    "MASS": {
      "Package": "MASS",
      "Version": "7.3-54",
      "Source": "Repository",
      "Repository": "CRAN",
      "Hash": "0e59129db205112e3963904db67fd0dc",
      "Requirements": []
    },
  ...
````

---

### `renv.lock` file

This __lockfile__ records all of the information about packages needed by your project.
* Version of package, where was it installed from, etc...

How does it know?
* `renv` scans all files in your project directory.
* Looks for `library`, `require`, or `package::function` syntax.
* Records those packages + their dependencies.

---

### `.Rprofile`

We also see `.Rprofile`, containing one line.

```{r, eval = FALSE}
source("renv/activate.R")
```

We have not talked about `.Rprofile`'s because they are generally antithetical to reproducible research.

When `R` starts, it searches for `.Rprofile` and runs what it finds.
* Use this to change various options, always load packages, etc...

In this case, whenever we start `R` in the project directory, we __activate__ our `R` environment.
* Telling `R` where packages for this project are saved, etc...
* Details are not too important.

---

### `renv` folder

You will also see a folder `example_project/renv/`.
* This folder contains your __project library__.

`renv` tries to be clever about installing packages.
* Already have a package installed elsewhere? `renv` will link to it.
* Otherwise, package is installed in `renv/library`.

???

Note that `renv/.gitignore` ensures packages not put under version control.

---

### Activating `renv`

`renv` will automatically be active in any `R` session that is run from the project directory.
* Recall the presence and function of `.Rprofile`.

Because our `Makefile` executes all code from the project directory, the switch to the project library should occur automatically.
* Use `.libPaths()` to see where `R` is looking for packages.

To activate in an `R` session run from elsewhere: `renv::activate('path/to/renv')`.

---

class: title-slide, center, inverse, middle
background-color: #487f84

.large[
`r fontawesome::fa("desktop", fill = "#ffffff")` .white[Showing project library is in use]
]

.large[
`r fontawesome::fa("code", fill = "#ffffff")` <a href="02_renv_hiv_project.zip" style="color:#ffffff" download>Code: `02_renv_hiv_project.zip`</a>
]

???

Demo: 
* `02_renv_hiv_project.zip`
* add a `print(.libPaths())` to `00_clean_data.R` to confirm project library is being used

